The function--sheaf dictionary for smooth commutative group schemes over finite fields with finitely generated geometric component group

### 2014-01-22 dictionary.tex
Just to see what it would like like, I've broken our paper into two parts: dictionary and geometrization. Please let me know what you think.

% !TEX encoding = UTF-8 Unicode
\documentclass{compositio}
\pdfoutput=1

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[A function--sheaf dictionary for local tori]{A function--sheaf dictionary for algebraic tori over local fields}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
\classification{14L15 (primary), 11G25, 14F05, 14D24, 11S31 (secondary)}
\keywords{quasicharacters, quasicharacter sheaves, algebraic tori, Greenberg transforms, N\'eron models, local fields, group schemes}

@@

\begin{abstract}
  We introduce and study the category of quasicharacter sheaves on smooth,
  commutative group schemes $G$ locally of finite type over finite fields
  $\Fq$. Assuming that the geometric component group of $G$
  is finitely-generated, we show that the group of isomorphism classes
  of quasicharacter sheaves on $G$ is canonically isomorphic to the group
  of characters of $G(\Fq)$. Then, for any non-archimedean local field $K$
  and any algebraic torus $T$ over $K$, we use this result to produce a functorial
  isomorphism between quasicharacters of $T(K)$ and isomorphism
  classes of quasicharacter sheaves on the Greenberg transform of the Néron
  model of $T$.
  \end{abstract}

@@

In this paper we find a dictionary between quasicharacters of algebraic tori over local fields and quasicharacter sheaves. 
What is a quasicharacter sheaf?
Let $G$ be a smooth, commutative group scheme locally of finite
type over a finite field $\Fq$; $G$ need not be reductive, connected, or of
finite type over $\Fq$. Fix a prime $\ell$ invertible in $\Fq$
and write $m : \GxG \to G$ for the multiplication morphism,
$\bG$ for the base change of $G$ to $\bFq$ and
$\Frob{G} : \bG \to \bG$ for the Frobenius automorphism.
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms;
$\mu$ and $\phi$ are required to satisfy natural conditions explained below.

In Part 1 of this paper we study the category $\QC(G)$ of quasicharacter sheaves on $G$.
Since $\QC(G)$ is a rigid monoidal category
the set of isomorphism classes $\QCiso{G}$ forms a group.
Motivated by the function--sheaf dictionary for commutative, connected
algebraic groups \cite{deligne:SGA4.5}*{sommes trig.}, 
we explore the trace of Frobenius morphism
\[
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\]
and show in Theorem~\ref{thm:snake} that it is an isomorphism, 
assuming that the geometric component group of $G$ is finitely generated.

In Part 2 we use Theorem~\ref{thm:snake} to make a function--sheaf dictionary
for algebraic tori over local fields.  Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fibre $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $n$, let $\GN{T}_n$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^n}$; the commutative
group scheme $\GN{T}_n$ plays the r\^{o}le of $G$, above.  Theorem~\ref{thm:application}
allows us to translate characters $T(K)$ of depth less than $n$ into quasicharacter sheaves on $\GN{T}_n$:
\[
\QCiso{\GN{T}_n} 
\iso
\Hom_{< n}(T(K),\EEx).
\]  
We also describe a pro-group scheme $\GN{T}$ 
simultaneously geometrizing all quasicharacters ({\it i.e.}, finite-depth characters) of $T(K)$.

While we have not yet fully explored the applications of such a geometrization, we want to highlight
some of the potential benefits of Theorem~\ref{thm:application} here. 
Characters on $T$ have limited
functoriality: they can be pulled back along maps to $T$, but it is less clear what to do with them
in the presence of maps from $T$ to other groups.  Sheaves, on the other hand, come
equipped with a host of functors, including pushforward. 
We hope to use this to our advantage as we work toward an adaptation of 
Lusztig's theory of characters sheaves to the study of characters of 
admissible representations of reductive $p$-adic groups that extends beyond the case of $p$-adic tori.

DeBacker and Reeder have recently applied Deligne-Lusztig 
representations to part of the depth-zero part of the
local Langlands correspondence \cite{reeder-debacker:09a}.
Deligne-Lusztig theory also plays an important role in 
Waldspurger's work on nilpotent distributions \cite{waldspurger:01}.
After developing Deligne-Lusztig representations, 
the next step Lusztig took in studying representations
of finite groups of Lie type was to introduce character sheaves for reductive groups over $\Fq$.
We hope that our quasicharacter sheaves 
can provide a stepping stone to a similar theory for
representations of $p$-adic groups, especially
in the positive-depth case.

The scope of the Langlands program has grown explosively in the past several decades, and
different parts of it apply to different kinds of representations.  In particular, the classic local Langlands
correspondence addresses complex (or $\ell$-adic) representations of real and $p$-adic groups, while
the geometric Langlands program aims at representations of groups in characteristic $p$.  The fact that
characteristic $p$ group schemes arise as the space underlying quasicharacter sheaves, regardless of
the characteristic of $K$, provides a potential bridge allowing the methods of the geometric Langlands
program to be applied to the classic context.

Finally, we are tantalized by the possibility of providing an alternate perspective on local class field theory
through quasicharacter sheaves.  The local Langlands correspondence for tori provides an isomorphism
$
\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \Hom(T(K), \EEx)
$. 
A geometric description of the resulting isomorphism $\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \QCiso{\GN{T}}$
would yield a different description of the Artin reciprocity map.

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}*{sommes trig.} and many people have since referred to such objects
as character sheaves; see for example, \cite{kamgarpour:09a}. 
Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the function--sheaf dictionary to the case of certain commutative
group schemes which are not necessarily connected or finite type.  The main use
of the term character sheaf is of course due to Lusztig. It is applied to certain perverse
sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
In Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} closes out Part \ref{part1} with the proof of the main theorem: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

In Section~\ref{sec:GN} we review the Greenberg transform of N\'eron models of tori, which produces
group schemes over $\Fq$ from algebraic tori $T$ over $K$.
Section~\ref{sec:bdchar} then uses Theorem~\ref{thm:snake} to give a geometrization of characters of $T$
of bounded depth.
We give a reinterpretation of this result in Section~\ref{sec:quasichar}, using a limit process to simultaneously
geometrize all quasicharacters of $T$.  Sections~\ref{sec:bdchar}, \ref{sec:quasichar} and \ref{sec:wrK} also include
functoriality results that we hope to use in a future discussion of 
local class field theory.
In Section~\ref{sec:transfer} we discuss a transfer principle which allows
a geometric transfer of characters between characteristic $p$ tori and characteristic $0$ tori.
We close the paper with Section~\ref{sec:reciprocity}, in which we suggest that quasicharacter
sheaves may play a key role in a geometric and categorical form of the local Langlands
correspondence for algebraic tori over local fields.

\begin{acknowledgements}

We would like to thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable. 
We also thank Alessandra Bertrapelle and Cristian Gonz\'ales-Avil\'es for disabusing us of a misapprehension concerning the Greenberg realization functor and for pointing us to a proof of Lemma~\ref{lemma:wrK}.

@@

\end{acknowledgements}

\part{Smooth commutative group schemes over finite fields} \label{part1}

@@

\section{The function--sheaf dictionary for non-connected group schemes}
\label{sec:snake}

@@

\part{Application to quasicharacters of algebraic tori over local fields}

Recall that $\Fq$ is a finite field.
Let $K$ be a non-archimedean local field with residue field $\Fq$.
We place no restriction on the characteristic of $K$. 
We write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$. 
Let $T$ be an algebraic torus over $K$; the filtrations defined by Moy-Prasad and Yu \cite{yu:03a}*{\S 5}
allow us to define the depth of character $\chi$ as the infimum of the $r$ so that $\chi$ vanishes on $T(K)_r$.
We write $\Hom_{< n}(T(K),\EEx)$ for the group of characters of $T(K)$ depth less than $n$ 
and $\Hom_{\text{qc}}(T(K),\EEx)$ for the group of characters of $T(K)$ of arbitrary finite depth;
we call these quasicharacters of $T(K)$, following \cite{tate:67a}*{2.3}.

\section{Greenberg of N\'eron}
\label{sec:GN}

A N\'eron model for $T$ is a smooth group scheme $\TT$ over $\OK$ with generic fiber $T$
that satisfies the N\'eron mapping property: if $S$ is any smooth scheme over $\OK$ then every morphism
$S_\eta \to T$ over $K$ extends uniquely to a morphism $S \to \TT$.  N\'eron models exist for any torus
$T$ \cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch 10.1, Thm 6} and are unique up to isomorphism;
$\TT$ is separated, commutative, locally finite type over $\OK$ and satisfies $\TT(\OK) = T(K)$.

We will also need the Greenberg functor \citelist{\cite{greenberg:61} \cite{greenberg:63a}}:
a functor from the category of schemes locally of finite type 
over an Artinian local ring with perfect residue field $k$ 
to the category of schemes locally of finite type over $k$. 
This functor respects open immersions, closed immersions, \'etale morphisms, smooth morphisms and geometric components.
See
\citelist{
\cite{demazure-gabriel:GroupesAlgebriques}*{V, \S 4, no. 1} \cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch. 9, \S 6} 
\cite{nicaise-sebag:motivicSerre}*{\S 2.2}
\cite{bertrapelle-gonzales:Greenberg}*{\S 5}
}
for the definition and fundamental properties of the Greenberg functor as it is used here.

\begin{definition}\label{def:Grn}
For any local field $K$ with residue field $\Fq$ and for any positive integer $n$, let
\[
\Gr_n : \partop{\text{Schemes}}{\text{lft over $\OK$}} \to \partop{\text{Schemes}}{\text{lft over $\Fq$}}
\]
be the functor produced by precomposing pullback along $\OK \to \OK/\pK^n$ with the Greenberg functor.
\end{definition}

The surjective ring homomorphism $\OK/\pK^{n+1} \to \OK/\pK^n$ determines a
natural transformation $P_{n} : \Gr_{n+1} \to \Gr_n$ between additive functors.
For a given Néron model $\TT$ of $T$, set
\begin{align*}
 \GN{T}_n &\ceq \Gr_n(\TT), \\
 f_n &\ceq P_n(\TT).
\end{align*}

We list some fundamental properties of $f_n$ and $\GN{T}_n$ for further use:
 \begin{enumerate}
  \labitem{(GN.0)}{truncation}
  $f_n : \GN{T}_{n+1} \to \GN{T}_n$ is a surjective morphism of group schemes over $\Fq$
  \citelist{\cite{bosch-lutkebohmert-reynaud:NeronModels}*{p. 276} \cite{bertrapelle-gonzales:Greenberg}*{Prop. 7.2}}.
  We remark that $f_n : \GN{T}_{n+1} \to \GN{T}_n$ is not induced from a morphism of Néron models.
  \labitem{(GN.1)}{smooth} 
  $\GN{T}_n$ is a $\QC$-scheme. 
  Note that $\underline{T}$ is a smooth group scheme over $\OK$, by construction. 
  Since smoothness is preserved under base change, 
  $\TT_n \ceq \underline{T} \times_\Spec{\OK} \Spec{\OK/\pK^n}$ 
  is a smooth group scheme over an Artinian ring with perfect residue field. 
  The Greenberg functor preserves smoothness, 
  so $\GN{T}_n$ is a smooth group scheme locally of finite type over $\Fq$. 
  The Greenberg functor takes the geometric identity component of $\TT_n$ 
  to the geometric identity component of $\GN{T}_n$.
  In fact, $\pi_0(\GN{T}_n) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$ and
  this group is finitely generated \cite{xarles:93a}*{2.18}.
  \labitem{(GN.2)}{points} $\GN{T}_n(\Fq) = \TT(\OK/\pK^n)$, canonically.
\end{enumerate}

\section{Application to quasicharacters of bounded depth} \label{sec:bdchar}

We now apply the results of Part~\ref{part1} to characters of $T(K)$.

\begin{theorem}\label{thm:application}
Let $n$ be a positive integer.
For every algebraic torus $T$ over $K$, the trace of Frobenius 
  \[
\trFrob{T,n}:  \QCiso{\GN{T}_n} \to \Hom_{< n}(T(K),\EEx)
  \]
is an isomomorphism. 
Moreover, the isomorphism $\trFrob{T,n}$ is a natural transformation
between the additive functors
\begin{enumerate}
\item[$F_1$:] $T \mapsto \QCiso{\GN{T}_n}$
\item[$F_2$:] $T \mapsto \Hom_{<n}(T(K),\EEx)$
\end{enumerate}
from the additive category of algebraic tori over $K$ to the category of abelian groups.
\end{theorem}

\begin{proof}
  By \ref{smooth}, we may apply Theorem~\ref{thm:snake} to
  $\GN{T}_n$ which provides a canonical isomorphism
  $\QCiso{\GN{T}_n} \iso \GN{T}_n(\Fq)^*$.
  By \ref{points}, $\GN{T}_n(\Fq) = \TT(\OK/\pK^n)$.
  Then we recall that the filtration of $T(K)$ used to
  define depth has the property that $\TT(\OK/\pK^n) \cong T(K) / T(K)_n$
  \cite{yu:03a}*{Prop 5.2}, so characters of $\GN{T}_n(\Fq)$ 
  are in bijection with characters of $T(K)$ vanishing on $T(K)_n$. 
  This shows that $\trFrob{T,n}$ is an isomorphism.

  The definition of $F_1$ depends 
  on the choice of a Néron model for every algebraic torus $T$.
  However, since Néron models are unique up to unique isomorphism,
  the isomorphism class of $\GN{T}_n$ is uniquely determined by $T$
  and the group of isomorphism classes of objects in $\QC(\GN{T}_n)$ 
  is uniquely determined by $T$. 
  To define $F_1$ on morphisms one uses the Néron mapping property.

  Finally, we show that $\trFrob{T,n}$ is functorial.
The first part of Proposition~\ref{prop:pullback} shows that $F_1$ is a functor,
while the second part shows that trace of Frobenius is a natural transformation
$t_{ - ,n} : F_1 \to F_2$. When further combined with Proposition~\ref{prop:product},
we see that $F_1$ is an additive functor and $t_{ - ,n}: F_1 \to F_2$ is a
natural transformation between additive functors.
\end{proof}

\section{A function--sheaf dictionary for tori over local fields} \label{sec:quasichar}

In this section we package the isomorphisms of Theorem~\ref{thm:application} into
a single dictionary.

\begin{definition}\label{def:objGN}
The \emph{Greenberg transform of the N\'eron model of $T$} 
is the pro-group scheme over $\Fq$ given by 
$\GN{T} \ceq \varprojlim_{n\in \ZZ_{>0}} \GN{T}_{n}$, 
with reference to the family of  surjective morphisms of 
$\Fq$-group schemes $f_n : \GN{T}_{n+1} \to \GN{T}_n$.
\end{definition}

For any  ordered pair of positive integers $n< m$ define 
$f_{n\leq m} : \GN{T}_{m} \to \GN{T}_{n}$ 
by $f_{n\leq m}\ceq f_n \circ f_{n+1} \circ \cdots \circ f_{m-1}$; 
also set $f_{n\leq n} \ceq \id_{\GN{T}_n}$.
Note that $f_{n\leq m}:  \GN{T}_{m} \to \GN{T}_{n}$ is a surjective morphism 
of group schemes over $\Fq$.
Let 
\[
f_{n\leq m}^* : \QC(\GN{T}_n) \to \QC(\GN{T}_{m})
\]
be the functor produced by applying Proposition~\ref{prop:pullback} to $f_{n\leq m}$.
We now define a category which allows us to extend the notion 
of quasicharacter sheaf to the pro-group scheme $\GN{T}$.

\begin{definition}\label{def:catGN}
An object in $\QC(\GN{T})$ is a triple $(n, \{\qcs{L}_i\}_{i \ge n}, \{\alpha_{i \le j}\}_{j \ge i \ge n})$, where $n$
is a positive integer, each $\qcs{L}_i$ is a quasicharacter sheaf on $\GN{T}_i$ and each
$\alpha_{i \le j} : \qcs{L}_j \to f^*_{i \le j} \qcs{L}_i$ is an isomorphism.  We require $\alpha_{i \le i}$ to be
the identity and that the $\alpha_{i \le j}$ are compatible with each other.  If $\qcs{L} \ceq (n, \{\qcs{L}_i\}, \{\alpha_{i \le j}\})$
and $\qcs{L}' \ceq (m, \{\qcs{L}'_i\}, \{\alpha'_{i \le j}\})$ are objects then $\Hom(\qcs{L}, \qcs{L}')$ is the set
of equivalence classes of pairs $(k, \{\beta_i\}_{i \ge k})$.  Here $k \ge m, n$ and the $\beta_i : \qcs{L}_i \to \qcs{L}'_i$ are
morphisms of quasicharacter sheaves so that
\[
\begin{tikzcd}
\qcs{L}_j \rar{\alpha_j} \dar{\beta_i} & f^*_{i \le j} \qcs{L}_i \dar{f_{i \le j}^*\beta_j} \\
\qcs{L}'_j \rar{\alpha_j} & f^*_{i \le j} \qcs{L}'_i
\end{tikzcd}
\]
commutes for all $j \ge i \ge k$.  We identify two such pairs $(k, \{\beta_i\})$ and $(l, \{\gamma_i\})$ if $\beta_i = \gamma_i$
for sufficiently large $i$.  Identities and composites are defined in the natural way.

If $n$ is a positive integer and $\qcs{L}_n \in \QC(\GN{T}_n)$ then we may define an object
\[
[\qcs{L}_n] \ceq (n, \{f_{n \le i} \qcs{L}_n\}_{i \ge n}, \{1_{\qcs{L}_j}\}_{j \ge i \ge n}).
\]
Every object in $\QC(\GN{T})$ is isomorphic to such a $[\qcs{L}_n]$.  Similarly, any morphism is equal to one of the form
\[
[\beta_k] \ceq (k, \{f_{k \le i}^* \beta_k\}_{i \ge k}).
\]

We may define a tensor product on $\QC(\GN{T})$ by setting
\[
 \qcs{L} \otimes \qcs{L}' \ceq (k, \{\qcs{L}_i \otimes \qcs{L}'_i\}, \{\alpha_{i \le j} \otimes \alpha'_{i \le j}\}),
\]
where $k = \max(m, n)$.
Similarly, the dual of an object $\qcs{L}$ is defined by 
\[
\qcs{L}^\vee \ceq (n, \{\qcs{L}_i^\vee\}, \{ \alpha_{i \le j}^\vee \}).
\]
We refer to objects in the rigid monoidal category 
$\QC(\GN{T})$ as \emph{quasicharacter sheaves on $\GN{T}$} 
or \emph{quasicharacter sheaves for $T$}.
\end{definition}

\begin{corollary}\label{cor:application}
For every algebraic torus $T$ over $K$, the trace of Frobenius 
  \[
\trFrob{T}:  \QCiso{\GN{T}} \to \Hom_{\text{qc}}(T(K),\EEx)
  \]
is an isomomorphism. 
Moreover, the isomorphism $\trFrob{T}$ is a natural transformation 
between the additive functors
\begin{enumerate}
\item[$F_{\QC}$:] $T \mapsto \QCiso{\GN{T}}$
\item[$F_{\text{qc}}$:] $T \mapsto \Hom_{\text{qc}}(T(K),\EEx)$
\end{enumerate}
from the additive category of algebraic tori over $K$ to the category of abelian groups.
\end{corollary}

\begin{proof}
We may define full and faithful functors $F_n : \QC(\GN{T}_n) \to \QC(\GN{T})$
for each $n$ by $F_n(\qcs{L}) \ceq [\qcs{L}_n]$ and $F_n(\alpha_n) \ceq [\alpha_n]$.
For positive integers $n \le m$ these functors fit together into a diagram
\[
\begin{tikzcd}
\ & & & \QC(\GN{T})\\
\arrow{urrr}{F_n} \QC(\GN{T}_n) \arrow[swap]{rr}{f_{n\leq m}^*} && \QC(\GN{T}_m) \arrow[swap]{ur}{F_m}. & 
\end{tikzcd}
\]
Passing to isomorphism classes of objects and invoking the 
universal property of direct limits gives a group isomorphism
$\theta: \varinjlim_{n\in \ZZ_{>0}} \QCiso{\GN{T}_n} \to \QCiso{\GN{T}}$
with inverse $[\qcs{L}]_{/\text{iso}} \mapsto [\qcs{L}_{/\text{iso}}]$.
Now use Theorem~\ref{thm:application}, Proposition~\ref{prop:pullback} 
and the fact that  $\Hom_{\text{qc}}(T(K),\EEx) = \varinjlim_{n\in \ZZ_{>0}} \GN{T}_n(k)^*$, 
where this direct limit is taken with respect to the family of group homomorphisms 
dual to $f_{n} : \GN{T}_{n+1}(\Fq) \to \GN{T}_n(\Fq)$.
\end{proof}

\section{Restriction of scalars over local fields}
\label{sec:wrK}

As indicated in the Introduction, we are tantalized 
by the idea that quasicharacter sheaves for $T$ 
may play a key role in a new geometric, categorial version 
of the local Langlands correspondence for algebraic tori 
over non-archimedean local fields. 
As with the local Langlands correspondence for algebraic tori over local fields, 
it should be possible to reduce the essential points of the argument to the case $T= \Gm{K}$,
provided that the category of quasicharacter sheaves behaves properly with regard to
Weil restriction, as in \cite{yu:09a}*{Thm 7.5 (2)}. 
The next lemma shows that the category of quasicharacter sheaves for 
algebraic tori does indeed behave properly with regard to Weil restriction.

\begin{lemma}\label{lemma:wrK}
Weil restriction is compatible with the Greenberg transform in the following sense: 
if $e$ is the ramification degree of $K'/K$ and $k'$ is the residue field of $K'$ then,
for every positive integer $n$,
\begin{equation}
\Gr_{n}\left( \Res_{\OKp/\OK}(\TTp)\right)
\iso
\Res_{k'/k} \Gr_{en}(\TTp).
\end{equation}
\end{lemma}

\begin{proof}
First, note that since $\Res_{\OKp/\OK}(\TTp)$ 
is a scheme over $\OK$, 
the Greenberg transform $\Gr_{n}\left( \Res_{\OKp/\OK}(\TTp)\right)$ 
is obtained by base change along $\OK \to \OK/\pK^n$ 
followed by the Greenberg functor over the Artinian ring $\OK/\pK^n$. 
On the other hand, since $\TTp$ is a scheme over $\OKp$, 
the Greenberg transform $\Gr_{en}(\TTp)$ is defined by base change along 
$\OKp \to \OKp/\pKp^{en}$ 
followed by the Greenberg functor over the Artinian ring $\OKp/\pKp^{en}$. 
Note also that 
\begin{equation}\label{nem}
\OKp \otimes_{\OK} \OK/\pK^n \iso \OKp/\pKp^{en}.
\end{equation}
Now Lemma~\ref{lemma:wrK} follows from \cite{bertrapelle-gonzales:Greenberg}*{Prop. 6.3} after observing that  $\TTp \times_\Spec{\OKp} \Spec{\OKp/\pKp^{en}}$ is admissible relative to $\Spec{\OKp/\pKp^{en}} \to \Spec{\OK/\pK^n}$, cf. \cite{bertrapelle-gonzales:Greenberg}*{Rem 2.9 (a)}.
\end{proof}

>>>>

% !TEX encoding = UTF-8 Unicode
%\documentclass{compositio}
\documentclass{amsart}
\pdfoutput=1

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[The function--sheaf dictionary]{The function--sheaf dictionary for smooth commutative group schemes over finite fields with finitely generated geometric component group}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
%\classification{14L15 (primary), 11G25, 14F05, 14D24, 11S31 (secondary)}
%\keywords{quasicharacters, quasicharacter sheaves, algebraic tori, Greenberg transforms, N\'eron models, local fields, group schemes}

@@

\begin{abstract}
  We introduce and study the category of quasicharacter sheaves on smooth,
  commutative group schemes $G$ locally of finite type over finite fields
  $\Fq$. Assuming that the geometric component group of $G$
  is finitely-generated, we show that the group of isomorphism classes
  of quasicharacter sheaves on $G$ is canonically isomorphic to the group
  of characters of $G(\Fq)$. 
   \end{abstract}

@@

The sheaf--function dictionary for connected commutative algebraic groups over finite fields translates characters into local systems  \cite{deligne:SGA4.5}*{sommes trig.}. More precisely, let $G$ be a commutative group scheme over a finite field $\Fq$, let $\ell$ be any prime invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character. Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$, the base change of $G$ to $\bFq$. While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism 
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The sheaf--function dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask if the sheaf--function dictionary can be generalized to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question.

In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $\bG$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notation of {\it quasicharacter sheaves on $G$}.

A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions. The main result of this paper is the following statement (see Theorem~\ref{thm:snake}, which says more): 
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and that the trace of Frobenius morphism provides an isomorphism
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx).
\end{equation}
In this way we find the geometrization and categorification of characters of $G(\Fq)$.

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to disconnected algebraic groups over finite fields. Consider the etale group scheme $G = \ZZ_{\Fq}$ over $\Fq$; this is a $\QC$-scheme.
Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general; for example, the $\ell$-adic character $\ZZ \to \EEx$ defined by $1 \mapsto 1+ \ell$ is not finite but bounded, which the character $\ZZ \to \EEx$ defined by $1 \mapsto \ell$ is unbounded.  As we show in the paper, local systems on $G$ cannot be used to geometrize either of these characters; however, quasicharacter sheaves do the job, handily. We remark that such characters arise when considering unramified quasicharacters of the $p$-adic group $\Qp^\times$. This example informs our choice of the term ``quasicharacter sheaf''. Indeed, in \cite{cunningham-roe:p-adic_tori} we use this paper to geometrize and categorify all quasicharacters of all algebraic tori over all local fields.

We also want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Some people refer to local systems satisfying \eqref{introbox} as as character sheaves; 
see for example, \cite{kamgarpour:09a}. 
As explained above, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig. 
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

We close the introduction with a more detailed description of the structure of this paper.
After setting some notation in Section~\ref{sec:QCschemes}, in Section~\ref{sec:category} we define quasicharacter sheaves and give an interpretation of
them as Weil sheaves with certain properties.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism,
one direction of the function--sheaf dictionary.
We define pullbacks and products of quasicharacter sheaves in Section~\ref{sec:pullback}, and use
them to prove a functoriality result on the trace of Frobenius.
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:bounded} and \ref{sec:finite} describe bounded and finite quasicharacter sheaves,
two subcategories of $\QC(G)$.  These alternate descriptions of quasicharacter sheaves play a key
role in the proof of Proposition~\ref{prop:restriction}.
In section \ref{sec:connected} we show that bounded and finite quasicharacter sheaves are the same
as standard quasicharacter sheaves when $G$ is connected, and that trace of Frobenius is an isomorphism.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and give a more concrete interpretation
of quasicharacter sheaves in this case.  Using a notion of global sections for quasicharacter sheaves,
we give isomorphisms between $\QCiso{G}$, $\Hh^1(\Weil{}, \Hom(G(\bFq), \EEx))$ and $\Hom(G(\Fq), \EEx)$.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
Section~\ref{sec:snake} presents the main theorem of this paper: trace of Frobenius
is an isomorphism for $G$ as long as the geometric component group is finitely generated.

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We would like to thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable. 
%We also thank Alessandra Bertrapelle and Cristian Gonz\'ales-Avil\'es for disabusing us of a misapprehension concerning the Greenberg realization functor and for pointing us to a proof of Lemma~\ref{lemma:wrK}.

@@

%\end{acknowledgements}

\section{Smooth commutative group schemes over finite fields}\label{sec:QCschemes}

@@

\section{The function--sheaf dictionary for disconnected group schemes}
\label{sec:snake}

@@
